---
title: "Chart Producing Hackathon: Visualizing the Happy Planet Index"
subtitle: "EPPS 6356 Data Visualization | Assignment 4"
author: " Hackathon Team:Nedret Sen-Cavus, Anugraha Samuel Gloria,Shao-Kang"
date: "October 2, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    fig-width: 10.5
    fig-height: 6.5
---
  
  ## Executive Overview
  
  This report completes Assignment 4 (48-Hour Chart Hackathon) for EPPS 6356 Data Visualization. We implement the core chart paradigms from Andrew Abela's *Chart Thought-Starter* (Comparison: Among Items branch) using the Happy Planet Index (HPI) dataset. All figures share a unified styling system defined in `R/theme.R`, adhere to Edward Tufte's data-ink maximization[cite: 4, 16], utilize color-blind safe palettes (Okabe & Ito 2008)[cite: 11], and encode comparisons along common spatial baselines (Cleveland & McGill 1984).

```{r}
#| label: setup
#| code-summary: "Load Dependencies, Theme, and Processed Datasets"

pal_okabe_ito <- c(
  "orange"        = "#E69F00",
  "sky_blue"      = "#56B4E9",
  "bluish_green"  = "#009E73",
  "yellow"        = "#F0E442",
  "blue"          = "#0072B2",
  "vermilion"     = "#D55E00",
  "reddish_purple"= "#CC79A7",
  "dark_gray"     = "#333333"
)

pal_continent <- c(
  "Africa"   = "#E69F00", # Orange
  "Americas" = "#56B4E9", # Sky Blue
  "Asia"     = "#009E73", # Bluish Green
  "Europe"   = "#0072B2", # Blue
  "Oceania"  = "#CC79A7"  # Reddish Purple
)

pal_rank_comparison <- c(
  "Top 20"    = "#009E73", # Greenish (high wellbeing / sustainable efficiency)
  "Bottom 20" = "#D55E00"  # Vermilion (low wellbeing / high footprint)
)

scale_fill_continent <- function(...) {
  scale_fill_manual(values = pal_continent, ...)
}

scale_color_continent <- function(...) {
  scale_color_manual(values = pal_continent, ...)
}

scale_fill_hackathon <- function(values = pal_okabe_ito, ...) {
  scale_fill_manual(values = values, ...)
}

scale_color_hackathon <- function(values = pal_okabe_ito, ...) {
  scale_color_manual(values = values, ...)
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(forcats)
  library(scales)
  library(gt)
  library(gtExtras)
})

hpi_clean     <- readRDS("../data/hpi_clean.rds")
df_chart1     <- readRDS("../data/hpi_chart1.rds")
df_chart2_raw <- readRDS("../data/hpi_chart2.rds")
df_chart3_raw <- readRDS("../data/hpi_chart3.rds")
```

---
  
  ## Chart 1: Variable-Width Column Chart
  
  ### Implementation (Comparison: Two Variables per Item)
  
  Chart 1 maps two continuous macro-demographic dimensions simultaneously using `geom_rect()`: continental population determines rectangle width ($W_i = xmax_i - xmin_i$), while population-weighted Happy Planet Index determines column height ($ymax_i$)[cite: 4, 5]. Interior polygon text has been eliminated to maximize the data-ink ratio and prevent label clipping across narrower columns (e.g., Oceania)[cite: 4, 5].

```{r}
#| label: chart-1-variable-width
#| code-summary: "Render Chart 1: geom_rect() Variable-Width Columns"
#| fig-width: 10.5
#| fig-height: 6.2

theme_hackathon <- function(base_size = 11,
                            base_family = "sans",
                            axis_slant = FALSE,
                            legend_pos = "top") {
  
  th <- theme_minimal(base_size = base_size, base_family = base_family) %+replace%
    theme(
      plot.title = element_text(
        size = rel(1.30),
        face = "bold",
        color = "#111111",
        hjust = 0,
        margin = margin(b = 6)
      ),
      plot.subtitle = element_text(
        size = rel(0.95),
        color = "#444444",
        hjust = 0,
        margin = margin(b = 12)
      ),
      plot.caption = element_text(
        size = rel(0.75),
        color = "#666666",
        hjust = 0,
        margin = margin(t = 12)
      ),
      
      axis.title.x = element_text(
        size = rel(0.90),
        face = "bold",
        color = "#222222",
        margin = margin(t = 8)
      ),
      axis.title.y = element_text(
        size = rel(0.90),
        face = "bold",
        color = "#222222",
        margin = margin(r = 8)
      ),
      axis.text = element_text(
        size = rel(0.82),
        color = "#333333"
      ),
      axis.line = element_line(
        color = "#777777",
        linewidth = 0.45
      ),
      axis.ticks = element_line(
        color = "#777777",
        linewidth = 0.40
      ),
      
      panel.grid.major = element_line(
        color = "#ECECEC",
        linewidth = 0.35
      ),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(
        fill = "#FFFFFF",
        color = NA
      ),
      plot.background = element_rect(
        fill = "#FFFFFF",
        color = NA
      ),
      
      legend.position = legend_pos,
      legend.title = element_text(
        size = rel(0.85),
        face = "bold",
        color = "#222222"
      ),
      legend.text = element_text(
        size = rel(0.80),
        color = "#333333"
      ),
      legend.key = element_blank(),
      legend.background = element_blank(),
      
      strip.background = element_rect(
        fill = "#F2F4F8",
        color = NA
      ),
      strip.text = element_text(
        size = rel(0.88),
        face = "bold",
        color = "#222222",
        margin = margin(t = 4, b = 4)
      ),
      
      plot.margin = margin(t = 16, r = 18, b = 14, l = 16),
      plot.title.position = "plot",
      plot.caption.position = "plot"
    )
  
  if (isTRUE(axis_slant)) {
    th <- th + theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
  }
  
  return(th)
}

p1 <- ggplot(df_chart1) +
  geom_rect(
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = ymax,
      fill = continent
    ),
    color = "#FFFFFF",
    linewidth = 0.7
  ) +
  geom_text(
    aes(
      x = xcenter,
      y = ymax + 1.6,
      label = sprintf("%.1f", ymax)
    ),
    color = "#222222",
    fontface = "bold",
    size = 3.4
  ) +
  scale_fill_continent(name = "Continent") +
  scale_x_continuous(
    name = "Cumulative Global Population (Millions)",
    expand = expansion(mult = c(0, 0.02)),
    breaks = seq(0, 5000, by = 1000),
    labels = comma_format(suffix = "M")
  ) +
  scale_y_continuous(
    name = "Population-Weighted Mean HPI (Score: 0–100)",
    expand = expansion(mult = c(0, 0.05)),
    limits = c(0, 75),
    breaks = seq(0, 70, by = 10)
  ) +
  labs(
    title = "The Americas Deliver Highest Wellbeing Efficiency Despite Lower Demographic Scale",
    subtitle = "Width represents total continent population; Column height reflects population-weighted Happy Planet Index.",
    caption = "Source: Happy Planet Index (2024 Edition) & UN Population Division | EPPS 6356 Hackathon"
  ) +
  theme_hackathon(legend_pos = "top") +
  theme(
    axis.title.y = element_text(
      angle = 90,
      vjust = 2,
      margin = margin(r = 10)
    ),
    legend.title = element_text(face = "bold", size = rel(0.85)),
    legend.margin = margin(b = 4)
  )

p1
```

### Analytical Takeaway
* **Demographic Scale vs. Ecological Efficiency:** While Asia represents over 60% of the sample's population ($\approx 3.9$ billion people), its aggregate HPI (44.6) is heavily constrained by industrial carbon throughput. In contrast, the Americas achieve the highest population-weighted efficiency (49.8), demonstrating that high human wellbeing can be sustained at lower ecological impact.

---

## Chart 2: Table with Embedded Distribution Bars

### Implementation (Comparison: Many Categories across Multiple Indicators)

Chart 2 presents 25 representative nations across development tiers. Built with `gt` and `gtExtras`, it embeds inline distribution bars directly into the data matrix, enabling pre-attentive length decoding along an aligned baseline while maintaining analytical precision[cite: 4, 10].

```{r}
#| label: chart-2-gt-table
#| code-summary: "Render Chart 2: gt + gtExtras Formatted Data Table"

df_table <- df_chart2_raw %>%
  arrange(desc(hpi)) %>%
  mutate(
    footprint_bar = (footprint / max(footprint)) * 100,
    hpi_bar = hpi
  ) %>%
  select(
    country, continent, hpi, hpi_bar,
    life_exp, wellbeing, footprint, footprint_bar, gdp_pc
  )

tbl_hpi <- df_table %>%
  gt() %>%
  tab_header(
    title = md("**High National Income Fails to Guarantee Ecological Efficiency Across Global Economies**"),
    subtitle = "Benchmarking 25 representative economies across life expectancy, ladder-of-life wellbeing, and ecological footprint"
  ) %>%
  cols_label(
    country = "Country",
    continent = "Continent",
    hpi = "HPI Score",
    hpi_bar = "Efficiency Bar",
    life_exp = "Life Expectancy (Years)",
    wellbeing = "Wellbeing (0–10)",
    footprint = "Footprint (gha/capita)",
    footprint_bar = "Pressure Bar",
    gdp_pc = "GDP per Capita (USD)"
  ) %>%
  cols_width(
    country ~ px(130),
    continent ~ px(95),
    hpi ~ px(70),
    hpi_bar ~ px(105),
    life_exp ~ px(100),
    wellbeing ~ px(80),
    footprint ~ px(95),
    footprint_bar ~ px(105),
    gdp_pc ~ px(120)
  ) %>%
  fmt_number(
    columns = c(hpi, life_exp, wellbeing),
    decimals = 1
  ) %>%
  fmt_number(
    columns = c(footprint),
    decimals = 2
  ) %>%
  fmt_currency(
    columns = c(gdp_pc),
    currency = "USD",
    decimals = 0
  ) %>%
  gt_plt_bar_pct(
    column = hpi_bar,
    scaled = TRUE,
    fill = "#0072B2",
    background = "#EAEAEA",
    height = 14
  ) %>%
  gt_plt_bar_pct(
    column = footprint_bar,
    scaled = TRUE,
    fill = "#D55E00",
    background = "#EAEAEA",
    height = 14
  ) %>%
  gt_badge(
    column = continent,
    palette = pal_continent
  ) %>%
  cols_align(
    align = "left",
    columns = c(country, continent)
  ) %>%
  cols_align(
    align = "center",
    columns = c(hpi, hpi_bar, footprint, footprint_bar)
  ) %>%
  cols_align(
    align = "right",
    columns = c(life_exp, wellbeing, gdp_pc)
  ) %>%
  tab_options(
    table.width = px(900),
    table.font.names = "sans-serif",
    table.font.size = px(13),
    heading.title.font.size = px(17),
    heading.title.font.weight = "bold",
    heading.subtitle.font.size = px(13),
    heading.align = "left",
    column_labels.font.size = px(12),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#F4F6F9",
    row.striping.include_table_body = TRUE,
    row.striping.background_color = "#FAFAFA",
    table.border.top.style = "solid",
    table.border.top.color = "#333333",
    table.border.top.width = px(2),
    table.border.bottom.style = "solid",
    table.border.bottom.color = "#333333",
    table.border.bottom.width = px(2),
    data_row.padding = px(5)
  ) %>%
  tab_source_note(
    source_note = md("Source: Happy Planet Index 2024; World Bank Development Indicators. Table compiled for EPPS 6356 Chart Hackathon.")
  )

tbl_hpi
```

### Analytical Takeaway
* **The Decoupling Paradox:** Economies with high GDP per capita (e.g., Luxembourg at \$116,300 and the United States at \$63,500) exhibit disproportionate ecological footprints (12.80 and 7.80 gha/capita), pulling their overall HPI scores below 45[cite: 4, 10]. Middle-income economies like Costa Rica (HPI 62.1) deliver comparable life expectancy with less than one-third of the ecological throughput[cite: 4, 10].

---

## Chart 3: Horizontal Bar Chart of Ranked Cohorts

### Implementation (Comparison: One Variable, Many Items)

Chart 3 benchmarks the Top 20 and Bottom 20 nations. Factor levels are ordered by continuous HPI score using `forcats::fct_reorder()`, eliminating alphabetical friction[cite: 4, 11]. The visual encoding uses an accessible, non-green Okabe-Ito color scheme (Okabe-Ito Blue `#0072B2` vs. Vermilion `#D55E00`)[cite: 11].

```{r}
#| label: chart-3-bar-chart
#| code-summary: "Render Chart 3: geom_col() + fct_reorder() Horizontal Ranked Bars"
#| fig-width: 11.5
#| fig-height: 7.8

df_chart3 <- df_chart3_raw %>%
  mutate(
    country_ordered = fct_reorder(country, hpi),
    group = factor(group, levels = c("Top 20", "Bottom 20"))
  )

p3 <- ggplot(df_chart3, aes(x = hpi, y = country_ordered, fill = group)) +
  geom_col(
    width = 0.72,
    show.legend = FALSE
  ) +
  geom_text(
    aes(label = sprintf("%.1f", hpi)),
    hjust = -0.20,
    size = 3.1,
    fontface = "bold",
    color = "#222222"
  ) +
  facet_wrap(
    ~ group,
    scales = "free_y",
    ncol = 2
  ) +
  scale_fill_manual(
    values = c(
      "Top 20"    = "#0072B2",  # Okabe-Ito Blue
      "Bottom 20" = "#D55E00"   # Okabe-Ito Vermilion
    )
  ) +
  scale_x_continuous(
    name = "Happy Planet Index Score (0–100 Scale)",
    limits = c(0, 74),
    breaks = seq(0, 70, by = 10),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_y_discrete(name = "Country") +
  labs(
    title = "Latin American Nations Lead Sustainable Efficiency While Fossil Rentiers and Fragile States Lag",
    subtitle = "Comparing the 20 highest- and 20 lowest-ranked global economies on the 2024 Happy Planet Index.",
    caption = "Source: Happy Planet Index (2024 Edition) | EPPS 6356 Hackathon"
  ) +
  theme_hackathon(legend_pos = "none") +
  theme(
    axis.text.y = element_text(size = rel(0.78), color = "#222222"),
    panel.spacing = unit(1.5, "lines"),
    strip.text = element_text(size = rel(0.95), face = "bold")
  )

p3
```

### Analytical Takeaway
* **Structural Distribution Divergence:** The Top 20 cohort is dominated by Latin American nations and high-efficiency island nations (e.g., Vanuatu)[cite: 4, 11]. The Bottom 20 features carbon-intensive fossil fuel exporters (e.g., Qatar, UAE, Mongolia) alongside developing economies with severe health infrastructure constraints (e.g., Central African Republic), highlighting two distinct drivers of low sustainable efficiency[cite: 4, 11].

---

## Computational Environment and Session Info

```{r}
#| label: session-info
#| code-summary: "Display sessionInfo() for Reproducibility"

sessionInfo()
```