---
title: "Chart Producing Hackathon: Visualizing the Happy Planet Index"
subtitle: "EPPS 6356 Data Visualization | Assignment 4"
author: " Hackathon Team:Nedret Sen-Cavus, Anugraha Samuel Gloria & Shao-Kang"
date: "October 2, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    fig-width: 10.5
    fig-height: 6.5
---

## Project Overview

For Assignment 4 in Dr. Karl Ho’s EPPS 6356 Data Visualization course, our objective was to operationalize Andrew Abela’s *Chart Thought-Starter* taxonomy using empirical data from the 2024 Happy Planet Index (HPI). The assignment required building a fully reproducible, self-contained repository in R, deploying the resulting figures to GitHub Pages, and demonstrating clear collaborative ownership.

As team coordinator, I designed the foundation for the team before any visualization code was drafted. In line with Dr. Ho’s recommendation to settle design foundations early, I initialized the remote GitHub repository. I then developed, debugged, and refined Charts 1, 2, and 3, and established the branch architecture to delegate Chart 4 to my teammates.

This report documents the architectural setup, mathematical boundaries, visual design theories, debugging challenges, and substantive findings that went into completing this sprint. All figures share a unified styling system, adhere to Edward Tufte's data-ink maximization, utilize color-blind safe palettes (Okabe & Ito 2008), and encode comparisons along common spatial baselines (Cleveland & McGill 1984).

## Visualization Engineering & Iteration

```{r}
#| label: setup
#| code-summary: "Load Dependencies, Theme, and Processed Datasets"
#| warning: false
#| message: false

suppressWarnings(suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(forcats)
  library(scales)
  library(gt)
  library(gtExtras)
}))
#| label: setup
#| code-summary: "Load Dependencies, Theme, and Processed Datasets"

pal_okabe_ito <- c(
  "orange"        = "#E69F00",
  "sky_blue"      = "#56B4E9",
  "bluish_green"  = "#009E73",
  "yellow"        = "#F0E442",
  "blue"          = "#0072B2",
  "vermilion"     = "#D55E00",
  "reddish_purple"= "#CC79A7",
  "dark_gray"     = "#333333"
)

pal_continent <- c(
  "Africa"   = "#E69F00", # Orange
  "Americas" = "#56B4E9", # Sky Blue
  "Asia"     = "#009E73", # Bluish Green
  "Europe"   = "#0072B2", # Blue
  "Oceania"  = "#CC79A7"  # Reddish Purple
)

pal_rank_comparison <- c(
  "Top 20"    = "#009E73", # Greenish (high wellbeing / sustainable efficiency)
  "Bottom 20" = "#D55E00"  # Vermilion (low wellbeing / high footprint)
)

scale_fill_continent <- function(...) {
  scale_fill_manual(values = pal_continent, ...)
}

scale_color_continent <- function(...) {
  scale_color_manual(values = pal_continent, ...)
}

scale_fill_hackathon <- function(values = pal_okabe_ito, ...) {
  scale_fill_manual(values = values, ...)
}

scale_color_hackathon <- function(values = pal_okabe_ito, ...) {
  scale_color_manual(values = values, ...)
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(forcats)
  library(scales)
  library(gt)
  library(gtExtras)
})

hpi_clean     <- readRDS("../data/hpi_clean.rds")
df_chart1     <- readRDS("../data/hpi_chart1.rds")
df_chart2_raw <- readRDS("../data/hpi_chart2.rds")
df_chart3_raw <- readRDS("../data/hpi_chart3.rds")
```

------------------------------------------------------------------------

## Chart 1: Variable-Width Column Chart

### Implementation (Comparison: Two Variables per Item)

Chart 1 maps two continuous macro-demographic dimensions simultaneously using `geom_rect()`: continental population determines rectangle width ($W_i = xmax_i - xmin_i$), while population-weighted Happy Planet Index determines column height ($ymax_i$). Interior polygon text has been eliminated to maximize the data-ink ratio and prevent label clipping across narrower columns (e.g., Oceania).

```{r}
#| label: chart-1-variable-width
#| code-summary: "Render Chart 1: geom_rect() Variable-Width Columns"
#| fig-width: 10.5
#| fig-height: 6.2

theme_hackathon <- function(base_size = 11,
                            base_family = "sans",
                            axis_slant = FALSE,
                            legend_pos = "top") {
  
  th <- theme_minimal(base_size = base_size, base_family = base_family) %+replace%
    theme(
      plot.title = element_text(
        size = rel(1.30),
        face = "bold",
        color = "#111111",
        hjust = 0,
        margin = margin(b = 6)
      ),
      plot.subtitle = element_text(
        size = rel(0.95),
        color = "#444444",
        hjust = 0,
        margin = margin(b = 12)
      ),
      plot.caption = element_text(
        size = rel(0.75),
        color = "#666666",
        hjust = 0,
        margin = margin(t = 12)
      ),
      
      axis.title.x = element_text(
        size = rel(0.90),
        face = "bold",
        color = "#222222",
        margin = margin(t = 8)
      ),
      axis.title.y = element_text(
        size = rel(0.90),
        face = "bold",
        color = "#222222",
        margin = margin(r = 8)
      ),
      axis.text = element_text(
        size = rel(0.82),
        color = "#333333"
      ),
      axis.line = element_line(
        color = "#777777",
        linewidth = 0.45
      ),
      axis.ticks = element_line(
        color = "#777777",
        linewidth = 0.40
      ),
      
      panel.grid.major = element_line(
        color = "#ECECEC",
        linewidth = 0.35
      ),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(
        fill = "#FFFFFF",
        color = NA
      ),
      plot.background = element_rect(
        fill = "#FFFFFF",
        color = NA
      ),
      
      legend.position = legend_pos,
      legend.title = element_text(
        size = rel(0.85),
        face = "bold",
        color = "#222222"
      ),
      legend.text = element_text(
        size = rel(0.80),
        color = "#333333"
      ),
      legend.key = element_blank(),
      legend.background = element_blank(),
      
      strip.background = element_rect(
        fill = "#F2F4F8",
        color = NA
      ),
      strip.text = element_text(
        size = rel(0.88),
        face = "bold",
        color = "#222222",
        margin = margin(t = 4, b = 4)
      ),
      
      plot.margin = margin(t = 16, r = 18, b = 14, l = 16),
      plot.title.position = "plot",
      plot.caption.position = "plot"
    )
  
  if (isTRUE(axis_slant)) {
    th <- th + theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )
  }
  
  return(th)
}

p1 <- ggplot(df_chart1) +
  geom_rect(
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = ymax,
      fill = continent
    ),
    color = "#FFFFFF",
    linewidth = 0.7
  ) +
  geom_text(
    aes(
      x = xcenter,
      y = ymax + 1.6,
      label = sprintf("%.1f", ymax)
    ),
    color = "#222222",
    fontface = "bold",
    size = 3.4
  ) +
  scale_fill_continent(name = "Continent") +
  scale_x_continuous(
    name = "Cumulative Global Population (Millions)",
    expand = expansion(mult = c(0, 0.02)),
    breaks = seq(0, 5000, by = 1000),
    labels = comma_format(suffix = "M")
  ) +
  scale_y_continuous(
    name = "Population-Weighted Mean HPI (Score: 0–100)",
    expand = expansion(mult = c(0, 0.05)),
    limits = c(0, 75),
    breaks = seq(0, 70, by = 10)
  ) +
  labs(
    title = "The Americas Deliver Highest Wellbeing Efficiency Despite Lower Demographic Scale",
    subtitle = "Width represents total continent population; Column height reflects population-weighted Happy Planet Index.",
    caption = "Source: Happy Planet Index (2024 Edition) & UN Population Division | EPPS 6356 Hackathon"
  ) +
  theme_hackathon(legend_pos = "top") +
  theme(
    axis.title.y = element_text(
      angle = 90,
      vjust = 2,
      margin = margin(r = 10)
    ),
    legend.title = element_text(face = "bold", size = rel(0.85)),
    legend.margin = margin(b = 4)
  )

p1
```

### Analytical Takeaway

- **The Population Weighting Effect:** Looking strictly at national averages, Latin American nations post the highest individual HPI scores globally. However, when weighting by population, **Europe leads all continents with a score of 49.4** across roughly 750 million people. The Americas place second at **46.7**; while countries like Costa Rica (62.1) and Colombia (60.2) demonstrate world-class efficiency, the region's aggregate score is pulled down by the demographic mass and heavy ecological footprint of the United States (37.4 across 333 million people).
- **Demographic Mass vs. Sustainable Return:** Asia spans the largest horizontal share of the chart—representing nearly 4 billion people in this sample—yet achieves an intermediate efficiency score of **44.9**. Oceania maintains a solid **46.0**, though its demographic footprint is a narrow sliver on the far right (roughly 32 million people). Africa records the lowest regional weighted score at **36.6** across roughly 880 million people, constrained primarily by systemic deficits in life expectancy and basic public health infrastructure rather than resource overconsumption.

------------------------------------------------------------------------

## Chart 2: Table with Embedded Distribution Bars

### Implementation (Comparison: Many Categories across Multiple Indicators)

Chart 2 presents 25 representative nations across development tiers. Built with `gt` and `gtExtras`, it embeds inline distribution bars directly into the data matrix, enabling pre-attentive length decoding along an aligned baseline while maintaining analytical precision.

```{r}
#| label: chart-2-gt-table
#| code-summary: "Render Chart 2: gt + gtExtras Formatted Data Table"

df_table <- df_chart2_raw %>%
  arrange(desc(hpi)) %>%
  mutate(
    footprint_bar = (footprint / max(footprint)) * 100,
    hpi_bar = hpi
  ) %>%
  select(
    country, continent, hpi, hpi_bar,
    life_exp, wellbeing, footprint, footprint_bar, gdp_pc
  )

tbl_hpi <- df_table %>%
  gt() %>%
  tab_header(
    title = md("**High National Income Fails to Guarantee Ecological Efficiency Across Global Economies**"),
    subtitle = "Benchmarking 25 representative economies across life expectancy, ladder-of-life wellbeing, and ecological footprint"
  ) %>%
  cols_label(
    country = "Country",
    continent = "Continent",
    hpi = "HPI Score",
    hpi_bar = "Efficiency Bar",
    life_exp = "Life Expectancy (Years)",
    wellbeing = "Wellbeing (0–10)",
    footprint = "Footprint (gha/capita)",
    footprint_bar = "Pressure Bar",
    gdp_pc = "GDP per Capita (USD)"
  ) %>%
  cols_width(
    country ~ px(130),
    continent ~ px(95),
    hpi ~ px(70),
    hpi_bar ~ px(105),
    life_exp ~ px(100),
    wellbeing ~ px(80),
    footprint ~ px(95),
    footprint_bar ~ px(105),
    gdp_pc ~ px(120)
  ) %>%
  fmt_number(
    columns = c(hpi, life_exp, wellbeing),
    decimals = 1
  ) %>%
  fmt_number(
    columns = c(footprint),
    decimals = 2
  ) %>%
  fmt_currency(
    columns = c(gdp_pc),
    currency = "USD",
    decimals = 0
  ) %>%
  gt_plt_bar_pct(
    column = hpi_bar,
    scaled = TRUE,
    fill = "#0072B2",
    background = "#EAEAEA",
    height = 14
  ) %>%
  gt_plt_bar_pct(
    column = footprint_bar,
    scaled = TRUE,
    fill = "#D55E00",
    background = "#EAEAEA",
    height = 14
  ) %>%
  gt_badge(
    column = continent,
    palette = pal_continent
  ) %>%
  cols_align(
    align = "left",
    columns = c(country, continent)
  ) %>%
  cols_align(
    align = "center",
    columns = c(hpi, hpi_bar, footprint, footprint_bar)
  ) %>%
  cols_align(
    align = "right",
    columns = c(life_exp, wellbeing, gdp_pc)
  ) %>%
  tab_options(
    table.width = px(900),
    table.font.names = "sans-serif",
    table.font.size = px(13),
    heading.title.font.size = px(17),
    heading.title.font.weight = "bold",
    heading.subtitle.font.size = px(13),
    heading.align = "left",
    column_labels.font.size = px(12),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#F4F6F9",
    row.striping.include_table_body = TRUE,
    row.striping.background_color = "#FAFAFA",
    table.border.top.style = "solid",
    table.border.top.color = "#333333",
    table.border.top.width = px(2),
    table.border.bottom.style = "solid",
    table.border.bottom.color = "#333333",
    table.border.bottom.width = px(2),
    data_row.padding = px(5)
  ) %>%
  tab_source_note(
    source_note = md("Source: Happy Planet Index 2024; World Bank Development Indicators. Table compiled for EPPS 6356 Chart Hackathon.")
  )

tbl_hpi
```

### Analytical Takeaway

- **The Income–Footprint Decoupling:** The inline pressure bars visually expose that high national wealth does not translate into sustainable efficiency. High-income economies like Luxembourg (\$116,300 GDP pc) and Qatar (\$93,500 GDP pc) achieve high life expectancy (82.6 and 79.3 years), but generate massive ecological footprints (12.80 and 14.30 gha/capita). These resource demands stretch their pressure bars to the maximum and drag their overall HPI scores down to **33.5** and **22.6**.
- **High Efficiency at Low Cost:** Middle- and low-income economies achieve comparable health and happiness outcomes at a fraction of the ecological throughput. Costa Rica (\$20,300 GDP pc) and Vanuatu (\$3,150 GDP pc) deliver life expectancies of 80.3 and 70.5 years with subjective wellbeing at 7.0/10, all while keeping footprints under 2.7 gha/capita. Their efficiency bars fill nearly two-thirds of the cell, demonstrating that human development does not require exponential carbon consumption.
- **Two Routes to Low Scores:** The bottom rows highlight two distinct failure mechanisms. In the United States (HPI 37.4) and Qatar (HPI 22.6), low efficiency stems from excessive consumption (footprints of 7.80 and 14.30 gha). In the Central African Republic (HPI 21.1), the low score is driven by acute humanitarian hardship—life expectancy drops to 53.9 years and wellbeing falls to 3.5, even though its footprint is minimal at 1.45 gha/capita.

------------------------------------------------------------------------

## Chart 3: Horizontal Bar Chart of Ranked Cohorts

### Implementation (Comparison: One Variable, Many Items)

Chart 3 compares the Top 20 and Bottom 20 HPI countries on one continuous horizontal scale. Countries are ordered by HPI using `forcats::fct_reorder()`, which removes alphabetical ordering and makes the ranking directly visible. Bar length and position encode HPI on a common baseline, while a color-blind-safe green–vermilion palette distinguishes the two cohorts. A separator line marks the boundary between the Top 20 and Bottom 20.

```{r}
#| label: chart-3-bar-chart
#| code-summary: "Render Chart 3: geom_col() + fct_reorder() Horizontal Ranked Bars"
#| fig-width: 11.5
#| fig-height: 7.8

source("Chart3_redesign.R", chdir = TRUE)

p3_redesign
```

### 

## Analytical Takeaway

- **Large spread across the ranking:** The continuous ordering reveals a substantial gap between the highest- and lowest-ranked countries. Costa Rica has the highest HPI score at **62.1**, while the Central African Republic has the lowest score at **21.1**, producing a difference of **41.0 points**. Displaying all 40 countries on one common scale makes this overall spread immediately visible and easier to compare than if the two cohorts were shown in separate panels.

- **Regional concentration among the highest-ranked countries:** Many of the highest HPI scores are concentrated in Latin America and the Caribbean. Costa Rica, Colombia, Ecuador, Panama, Jamaica, Guatemala, Honduras, Nicaragua, Peru, Mexico, and the Dominican Republic all appear in the Top 20. However, the Top 20 is not regionally uniform, as countries such as Switzerland, Norway, Finland, Spain, Japan, Vietnam, and the Philippines also rank highly.

- **The Bottom 20 is geographically diverse:** The lower end of the ranking includes countries from Africa, Europe, Asia, the Americas, and the Middle East. Examples include Senegal, Ghana, Kenya, Estonia, Bulgaria, Egypt, Uganda, the United States, Tanzania, Saudi Arabia, Ethiopia, Nigeria, Luxembourg, the United Arab Emirates, South Africa, Sierra Leone, Mongolia, Zimbabwe, Qatar, and the Central African Republic. This indicates that low HPI scores are not confined to a single region.

- **The visualization emphasizes comparison rather than explanation:** Chart 3 shows how countries differ in HPI score, but it does not by itself explain why those differences occur. Variables such as life expectancy, wellbeing, ecological footprint, and GDP per capita are present in the dataset, but they are not encoded in this chart. Therefore, the chart should be interpreted as a ranked comparison of HPI outcomes rather than as evidence of causal relationships.

------------------------------------------------------------------------

## Technical Troubleshooting & Debugging Log

During development, several technical issues required hands-on troubleshooting to keep the project moving:

- **Package Installation & System Permissions:** Running `install.packages("dplyr", "tidyr", ...)` initially threw directory errors because missing `c()` wrappers caused R to misinterpret package names as local paths and download repositories. In addition, Windows compilation prompts for Rtools appeared. I resolved this by setting `options(pkgType = "binary")` to pull pre-compiled binaries, avoiding unnecessary local compiler builds.
- **Quarto Path Resolution:** Early preview runs failed with `cannot open the connection` on `source("R/theme.R")` because the file was saved in an external folder. I wrote an inline script generator to save `R/theme.R` directly into the project root, satisfying the self-contained repository requirement.
- **YAML Parser Crashes:** Quarto threw `readBlockMapping2` errors due to a merged title/subtitle line and non-breaking space characters (`\u00A0`) in the YAML header. I rebuilt the front matter with standard two-space indentation and dedicated lines for every key.
- **Table Cell Clipping:** The gt table's right-aligned currency columns were getting clipped during headless rendering in Chrome. Setting fixed pixel widths on every column via `cols_width()` resolved the overflow.

For all these errors I asked Gemini- Flash Extended model. I gave my code and the warning error I got in the console to Gemini and asked why it happpened, and how I could fix it.

## Computational Environment and Session Info

```{r}
#| label: session-info
#| code-summary: "Display sessionInfo() for Reproducibility"

sessionInfo()
```
